---
title: "Lab #7"
author: "Julian Stanley"
date: "6 April 2018"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Introductory Matter
---------------

Ancova: numerical response, two explanatory: categorical and continuous

```{r}
x <- rnorm(100)
slopeA <- 2
interceptA <- 0
slopeB <- -1
interceptB <- 0

# Create a response variable as a function of treatment A
yA <- slopeA * x + interceptA + rnorm(100)

# Create a response variable as a function of treatment B
yB <- slopeB * x * interceptB + rnorm(100)

# Create a datafram with our yA and yB
myDat <- data.frame(y = c(yA, yB), x = c(x, x), trt = rep(c("A", "B"), 
                                                          each = length(x)))

str(myDat)

# Visualize data
plot(x, yA, col = "red")
points(x, yB, col = "blue")
abline(lm(yA ~ x), col = "red")
abline(lm(yB ~ x), col = "blue")

# Make a linear model object with interaction
mod <- lm(y ~ x * trt, data = myDat)

shapiro.test(mod$residuals)
qqnorm(mod$residuals)
qqline(mod$residuals, col = "red")

anova(mod)


```


### Challenge 1

```{r}
x <- rnorm(100)
slopeA <- 0
interceptA <- 0
slopeB <- 0
interceptB <- 0

# Create a response variable as a function of treatment A
yA <- slopeA * x + interceptA + rnorm(100)

# Create a response variable as a function of treatment B
yB <- slopeB * x * interceptB + rnorm(100)

# Create a datafram with our yA and yB
myDat <- data.frame(y = c(yA, yB), x = c(x, x), trt = rep(c("A", "B"), 
                                                          each = length(x)))

str(myDat)

# Visualize data
plot(x, yA, col = "red")
points(x, yB, col = "blue")
abline(lm(yA ~ x), col = "red")
abline(lm(yB ~ x), col = "blue")

# Make a linear model object with interaction
mod <- lm(y ~ x * trt, data = myDat)

shapiro.test(mod$residuals)
qqnorm(mod$residuals)
qqline(mod$residuals, col = "red")

anova(mod)


```

### Challenge 2

```{r}
x <- rnorm(100)
slopeA <- 0
interceptA <- 0
slopeB <- 0
interceptB <- 2

# Create a response variable as a function of treatment A
yA <- slopeA * x + interceptA + rnorm(100)

# Create a response variable as a function of treatment B
yB <- slopeB * x * interceptB + rnorm(100)

# Create a datafram with our yA and yB
myDat <- data.frame(y = c(yA, yB), x = c(x, x), trt = rep(c("A", "B"), 
                                                          each = length(x)))

str(myDat)

# Visualize data
plot(x, yA, col = "red")
points(x, yB, col = "blue")
abline(lm(yA ~ x), col = "red")
abline(lm(yB ~ x), col = "blue")

# Make a linear model object with interaction
mod <- lm(y ~ x * trt, data = myDat)

shapiro.test(mod$residuals)
qqnorm(mod$residuals)
qqline(mod$residuals, col = "red")

anova(mod)


```

### Challenge 3

```{r}
x <- rnorm(100)
slopeA <- 2
interceptA <- 0
slopeB <- 2
interceptB <- 0


# Create a response variable as a function of treatment A
yA <- slopeA * x + interceptA + rnorm(100)

# Create a response variable as a function of treatment B
yB <- slopeB * x * interceptB + rnorm(100)

# Create a datafram with our yA and yB
myDat <- data.frame(y = c(yA, yB), x = c(x, x), trt = rep(c("A", "B"), 
                                                          each = length(x)))

str(myDat)

# Visualize data
plot(x, yA, col = "red")
points(x, yB, col = "blue")
abline(lm(yA ~ x), col = "red")
abline(lm(yB ~ x), col = "blue")

# Make a linear model object with interaction
mod <- lm(y ~ x * trt, data = myDat)

shapiro.test(mod$residuals)
qqnorm(mod$residuals)
qqline(mod$residuals, col = "red")

anova(mod)


```


### Challenge 4

```{r}
x <- rnorm(100)
slopeA <- 2
interceptA <- 4
slopeB <- 2
interceptB <- 0


# Create a response variable as a function of treatment A
yA <- slopeA * x + interceptA + rnorm(100)

# Create a response variable as a function of treatment B
yB <- slopeB * x * interceptB + rnorm(100)

# Create a datafram with our yA and yB
myDat <- data.frame(y = c(yA, yB), x = c(x, x), trt = rep(c("A", "B"), 
                                                          each = length(x)))

str(myDat)

# Visualize data
plot(x, yA, col = "red")
points(x, yB, col = "blue")
abline(lm(yA ~ x), col = "red")
abline(lm(yB ~ x), col = "blue")

# Make a linear model object with interaction
mod <- lm(y ~ x * trt, data = myDat)

shapiro.test(mod$residuals)
qqnorm(mod$residuals)
qqline(mod$residuals, col = "red")

anova(mod)


```

Main Assignment Task 1
----------------

### Part 1

```{r}
# Download data
d1 <- read.csv("http://faraway.neu.edu/data/lab6_dataset1.csv")

str(d1)
head(d1)
```
Data has two categorical treatments: d2 treatment or d1 treatment, and one numerical response variable: score

### Part 2

```{r}
# Visualize distribution
par(mfrow = c(2,2))
hist(d1$score, main = "Raw Data")
hist(log(d1$score), main = "Log Transformed Data")

qqnorm(d1$score, main = "QQ Raw Data")
qqline(d1$score)
qqnorm(log(d1$score), main = "QQ Log Transformed")
qqline(log(d1$score))

shapiro.test(d1$score)
shapiro.test(log(d1$score))
```
Data is greatly improved by a log transformation of the numerical response variable.

### Part 3


```{r}
interactionModel <- aov(log(d1$score) ~ drug1 * drug2, data = d1)
summary(interactionModel)
```

### Part 4

We conclude that the interaction is significant (p < 0.01).

### Part 5

```{r}
tukeyModel <- TukeyHSD(interactionModel)
par(mar = c(3,11,3,1))
plot(TukeyHSD(interactionModel), las = 1)
```


### Part 6

There is a significant difference between the treatment and control of drug 2

### Part 7
```{r}
se <- function(x) sqrt(var(x)/length(x))

print("Means:")
(means <- tapply(d1$score, list(d1$drug1, d1$drug2), FUN = mean))
print("Standard Errors:")
(ses <- tapply(d1$score, list(d1$drug1, d1$drug2), FUN = se))

require(multcompView)
(labels <- multcompLetters(tukeyModel$`drug1:drug2`[, "p adj"])$Letters)
labels <- labels[c("d1.ctrl:d2.ctrl", "d1.trt:d2.ctrl", "d1.ctrl:d2.trt", "d1.trt:d2.trt")]

bar <- barplot(means, beside = TRUE, legend.text = c("D1 Control", "D1 Treatment"), names = c("D2 Control", "D2 Treatment"), ylab = "Score", col = c("aquamarine3", "coral3"), ylim = c(0, 1200))

arrows(x0 = bar, y0 = c(means[1] - 2*ses[1], means[2] - 2*ses[2], means[3] - 2*ses[3], means[4] - 2*ses[4]) , x1 = bar, y1 = c(means[1] + 2*ses[1], means[2] + 2*ses[2], means[3] + 2*ses[3], means[4] + 2*ses[4]), angle = 90, code = 3)

text(x = bar, y = c(means[1] + 2*ses[1], means[2] + 2*ses[2], means[3] + 2*ses[3], means[4] + 2*ses[4]), labels, pos = 3)
```

### Part 8

```{r}
interaction.plot(d1$drug1, d1$drug2, response = d1$score)
``` 


### Part 9

The cocktail of both drug1 and drug2 is more effective than either drug by itself

Task 2
----------------

```{r}
d2 <- read.csv("http://faraway.neu.edu/data/lab6_dataset2.csv")

str(d2)
head(d2)
```

### Part 1

```{r}
hist(log(d2$score))
model2 <- aov(log(d2$score) ~ drug1 + drug2 + sex + drug1*drug2 + drug2*sex + drug1*drug2*sex, data = d2)
summary(model2)
```
Data is log-normal. Drug 1 has no effect, there is no interaction between drug 1 and sex. There is an effect of drug 2 and of sex, and an interaction between drug2 and sex, drug1 and drug2, and all three variables.

### Part 2

```{r}
par(mfrow = c(2,1))
interaction.plot(d2$drug1, d2$sex, response = d2$score)
interaction.plot(d2$drug2, d2$sex, response = d2$score)
```

### Part 3

Female and drug 1 treatment are antagonistic, female and drug 1 are synergistic.

### Part 4

```{r}
d2mean <- aggregate(score ~ drug1 + drug2 + sex, FUN = mean, data = d2)
d2se <- aggregate(score ~ drug1 + drug2 + sex, FUN = se, data = d2)

tukeyModel2 <- TukeyHSD(model2)

bar2 <- barplot(cbind(d2mean[1:4,4], cbind(d2mean[5:8, 4])), beside = TRUE, legend.text = c("Drug 1 Control + Drug 2 Control", "Drug 1 Treatment + Drug 2 Control", "Drug 1 Control + Drug 2 Treatment", "Drug 1 Treatment + Drug 2 Treatment"), args.legend = list(x = "topleft"), names = c("Female", "Male"), ylim = c(0, 2000))

arrows(x0 = bar2, y0 = d2mean$score - 2*d2se$score , x1 = bar2, y1 = d2mean$score + 2*d2se$score, angle = 90, code = 3)

require(multcompView)
(labels <- multcompLetters(tukeyModel2$`drug1:drug2:sex`[, "p adj"])$Letters)
labels <- labels[c("d1.ctrl:d2.ctrl:female", "d1.trt:d2.ctrl:female", "d1.ctrl:d2.trt:female", "d1.trt:d2.trt:female", "d1.ctrl:d2.ctrl:male", "d1.trt:d2.ctrl:male", "d1.ctrl:d2.trt:male", "d1.trt:d2.trt:male")]
text(x = bar2, y = d2mean$score + 2*d2se$score, labels, pos = 3)

```

### Task 5

There is no significant difference between treatment and no treatment in females.

In males, a combination therapy is the most effective treatment. 

