---
title: "Lab #3"
author: "Julian Stanley"
date: "23 February 2018"
output: html_document
---

```{r setup, include=FALSE}
# This is where you set global options (leave as-is)
knitr::opts_chunk$set(echo = TRUE)
```



### Introductory Material

```{r}
my.vars <- data.frame(a1 = 1:10, a2 = 11:20, b1 = 21:30, b2 = 31:40)
my.means <- sapply(my.vars, FUN = mean)
my.means

# Get the mean of the first and third columns
my.first.means <- sapply(my.vars[,c(1,3)], FUN = mean)
my.first.means

my.mat <- cbind(my.means[c(1,3)], my.means[c(2,4)])
my.mat

barplot(my.mat,beside = TRUE, names.arg = c("a1", "b1", "a2", "b2"))
```


### Exercise 1

```{r}
my.mat2 <- cbind(my.means[c(1,2)], my.means[c(3,4)])
my.mat2

# Function that computes 95 percent confidence intervals
conf95 <- function(x) {
  return(1.96*(sd(x) / sqrt(length(x))))
}

# upper conf interval
my.conf_upper = my.means + sapply(my.vars, FUN = conf95)
my.conf_lower = my.means - sapply(my.vars, FUN = conf95)

bar1 <- barplot(my.mat2,beside = TRUE, names.arg = c("a1", "a2", "b1", "b2"), ylim = c(0,40))
arrows(x0 = bar1, y0 = my.conf_lower , x1 = bar1, y1 = my.conf_upper, angle = 90, code = 3)

my.conf_upper
```



```{r}
my.sums <- colSums(my.vars)
```


```{r}
# Create some data
my.table <- cbind(c(5, 21), c(65, 56))
my.table
f.test <- fisher.test(my.table)
f.test

# Extract the p-value
f.test$p.value
```



Task 1
------------

### Problem 1
Setup: Fields with alternating GMO and non-GMO plants. Blind: neither the researcher running the experiment nor the person planting the seeds know which seeds are GMO and which are non-GMO.

Controls: Need a positive control. I would also plant some seeds with a known-low gemination rate and a known-high germination rate as positive and negative controls. 

Replication: Use multiple fields and as many seeds as possible

### Problem 2
```{r}
d1 <- read.csv("http://faraway.neu.edu/data/lab3_dataset1.csv")
```




### Problem 3
```{r}
colnames(d1)

# Seeds germinated vs seeds that failed to germinate in GMO and wild type

seed.means <- sapply(d1[2:5], FUN = sum)
seed.means

seeds.conf.lower <- seed.means - sapply(d1[2:5], FUN = conf95)

seeds.conf.upper <- seed.means +  sapply(d1[2:5], FUN = conf95)
```


### Problem 4
```{r}
seedsBar <- barplot(seed.means, names.arg = c("GMO Germ", "Wild Germ", "GMO NoGerm", "Wild NoGerm"), beside = FALSE, ylim = c(0,1000))
arrows(x0 = seedsBar, y0 = seeds.conf.lower , x1 = seedsBar, y1 = seeds.conf.upper, angle = 90, code = 3)
```


### Problem 5
It looks fantastic! GMO plants are germinating a lot more than wild plants, meethinks.

### Problem 6
$H_0$: Wild-type and GMO plants germinate at the same rate
$H_a$: Wild-type and GMO plants do not germinate at the same rate


### Problem 7
```{r}

seed.sum <- sapply(d1[2:5], FUN = sum)

seed.table <- cbind(c(as.numeric(seed.sum[1]), as.numeric(seed.sum[3])), c(as.numeric(seed.sum[2]), as.numeric(seed.sum[4])))
seed.table
seed.test <- fisher.test(seed.table)
seed.test

# Extract the p-value
seed.test$p.value
```

### Problem 8
Yes! p>1E-14. There is very likely a difference between wild-type and GMO plants

### Problem 9
```{r}
seed.test$estimate
```


### Problem 10
My research produced a GMO plant that is eleven-times more likely to germinate than normal the wild plants.


Task 2
----------------------

```{r}
# Download the dataset
d2 <- read.csv("http://faraway.neu.edu/data/lab3_dataset2.csv")
```


### Problem 1

$H_0$: GMO use does not change disease incidence
$H_a$: GMO use changes disease incidence

### Problem 2
```{r}
sapply(d2[2:5], FUN = sum)
disease.table <- cbind(c(506, 465), c(228, 260))
fisher.test(disease.table)
```

### Problem 3

There may be a small association between GMO and disease, but it is not yet significant (p > 0.05).

### Problem 4

```{r}
fin <- matrix(NA, nrow = nrow(d2), ncol = 1)

for (i in 1:nrow(d2)) {
  v <- d2[i, 2:5]
  v.table <- cbind(c(as.numeric(v[1]), as.numeric(v[3])), c(as.numeric(v[2]), as.numeric(v[4])))
  fin[i, 1] <- fisher.test(v.table)$p.value
}

rownames(fin) <- as.character(d2$countries)
colnames(fin) <- "P Values"
fin

barplot(fin[1:10,], horiz = FALSE, ylab = "P Value", xlab = "Countries")

```


### Problem 5

There is only a significant (p< 0.05) coorelation between GMO use and disease in India.

### Problem 6

I would recommend neither a ban nor moratorium. Proper controls were not used to generate this data, and therefore what we are seeing is a correlation, not a causation. 




