---
title: "Lab #2"
author: "Julian Stanley"
date: "16 February 2018"
output: html_document
---

```{r setup, include=FALSE}
# This is where you set global options (leave as-is)
knitr::opts_chunk$set(echo = TRUE)
```


Some Introductory Material
-------------------------

```{r}
a <- rnorm(100)
hist(a)

hist(a, main = "Title", xlab = "X label", ylab = "Y label")

my.hist <- hist(a)
my.hist
```

```{r}
my.hist$counts

# Change number to a character
b <- rnorm(1)

b

as.character(b)
```

```{r}
my.string <- as.character(b)
substr(my.string, start = 1, stop = 3)
```

```{r}
c <- 10:20
length(c)
```

```{r}
# Binominal test 
n <- 10
k <- 8
p <- 0.5

binom.test(x = k, n = n, p = p)
```

```{r}
# Chi-sqr test
observed <- 1:10
expected <- 1:10

chisq.test(x = observed, p = expected/sum(expected))
```

```{r}
bp <- barplot(observed, names.arg = c(1:10))
points(bp, expected)
```

Task 1
--------

### Problem 1

```{r}

# Probability of 3 out of 20 species contracting their range southward
dbinom(3, 21, p = 0.5)
```

### Problem 2
```{r}
# As extreme or more extreme than 3 out of 20; two tailed
pbinom(3, 21, p = 0.5) * 2
```

### Problem 3
```{r}
# As extreme or more extreme than 3 out of twenty equals probability of as extreme or more extreme than 18
pbinom(17, 21, p = 0.5, lower.tail = FALSE) *2 ==
pbinom(3, 21, p = 0.5, lower.tail = TRUE) *2
```


Task 2
---------------

### Problem 1

```{r}
d2012 <- read.csv("http://faraway.neu.edu/data/lab2_dataset1.csv")

# Use substr to extract first digit
first.digit <- substr(as.character(d2012$voted.democrat), start = 1, stop = 1)

# Convert the first digit to a number
first.digit <- as.numeric(first.digit)
```


### Problem 2
```{r}
# Use the seq() function to create a sequence of numbers
seq(0.5, 9.5, by = 1)
digit.hist <- hist(first.digit, breaks = seq(0.5, 9.5, by = 1), plot = FALSE)
head(digit.hist)

observed <- digit.hist$counts

hist(first.digit)$counts

digit.hist
```

### Problem 3

```{r}
# Benford's law (gives a proportion)
(expected <- log10(1 + 1/(1:9)))
# Expected count for each digit based on Benford's Law
(expected <- round(expected * length(first.digit))/sum(expected))
```

### Problem 4

```{r}
obPlot <- barplot(observed, ylim = c(0, 400))
points(obPlot, expected)
```


### Problem 5
It does not look consistent with Benford's law

### Problem 6

Null Hypothesis: The election counts conform with Benford's law

### Problem 7
```{r}
chisq.test(x = observed, p = expected/sum(expected))
```

### Problem 8
Fail to reject null hypothesis: the election counts conform with Benford's law

Task 3
--------------

### Problem 1

$$P(F|W) = 0.75$$

### Problem 2

$$P(F|\bar{W}) = 0.20$$



### Problem 3

$$P(W|F) =\frac{P(F|W) * P(W)}{P(F)}$$

### Problem 4

$$P(F) = P(F|W)*P(W) + P(F|\bar{W})*(1-P(W))$$

```{r}
# P(Win)
pW=0.5

# P(Favored|Win)
pF.W= 0.75

# P(Favored|Loss)
pF.L= 0.2

# P(Favored)
pF= (pF.W * pW) + (pF.L * (1-pW))

# P(Win|Favored)
pW.F= pF.W * pW / pF


```


### Problem 5

```{r}
# P(Win) is now a vector between zero and one
pWvals=seq(0, 1, length=100)


# P(Favored|Win) is still the same
pF.W= 0.75

# P(Favored|Loss) is still the same
pF.L= 0.2

# P(Favored) formula is still the same, just pW --> pwVals, pL --> pLvals
pF = (pF.W * pWvals) + (pF.L * (1-pWvals))

# P(Win| Favored) formula is still the same, pW --> pwVals, pL --> pLvals
pW.Fvals <- pF.W * pWvals / pF


```



### Problem 6

```{r}
plot(x = pWvals, y = pW.Fvals, abline(b = 1, a = 0),
     col = "blue", xlab = "Probability of Winning", ylab = "Probability of winning, given favored in polls", main = 
      "Change in P(W|F) with the change in P(W)")
```


When the blue line is below the black line,$P(W|F) < P(F)$, whereas when the blue line is above the black line, $P(W|F) > P(F)$. In this case, the chances of winning, given that you are favored is almost always higher than the probability of winning. That difference is most dramatic, however, around when you have around a 40% chance of winning and are favored. 


Task 4
----------------

### Problem 1

```{r}
hist(d2012$population.size, main = "2012 Pop Size", xlab = "population size", ylab = "frequency")
```

The population frequency is left-skewed


### Problem 2
```{r}
# Compute the skewness of this vector
skewness <- function(x) {
  # X: Vector numbers
  mean <-  mean(x)
  sd = sd(x)
  
  return (sum(((x - mean) / sd)^3))
}
```

### Problem 3
```{r}
skewness(d2012$population.size)
mean(d2012$population.size)
median(d2012$population.size)
```


### Problem 4

```{r}
hist(d2012$voted.democrat / d2012$voted.republican, main = "Ratio of voting Democrat/Republican, by district", xlab = "Ratio Dem/Rep", ylab = "frequency")
```

```{r}
dem = d2012$voted.democrat
rep = d2012$voted.republican

hist(dem, col = adjustcolor('blue', alpha = 0.3), main = "Republican VS Democrat voting distributions", xlab = "Votes", ylab = "frequency")
par(new = TRUE)
hist(rep, col = adjustcolor('red', alpha = 0.3), main = "", xlab = "", ylab = "", axes = FALSE)

legend("topright", legend = c("Democrat", "Republican"), col = c("blue", "red"), pch = c(19,19))
```


The distributions of democrat and republican votes have a similar distributions
