---
title: "Lab #4"
author: "Julian Stanley"
date: "03 March 2018"
output: html_document
---

```{r setup, include=FALSE}
# This is where you set global options (leave as-is)
knitr::opts_chunk$set(echo = TRUE)
```


Introductory Matter
-------------

```{r}
x <- rnorm(100000)

## make a histogram
hist(x)

## Make a QQ plot
qqnorm(x)

## Add a theoretical line
qqline(x, col = "red")
```


```{r}
# Negative control: different distribution

y <- rlnorm(1000)

# Histogram
hist(y)

## QQ plot
qqnorm(y)

## Still can add a line
qqline(y, col = "red")

# Log transform
log_y <- log10(y)
hist(log_y)
qqnorm(log_y)
qqline(log_y, col = "red")
```



Let's do some T tests


Task 1
-------------

```{r}
# Download the data
d1 <- read.csv("https://nofile.io/f/jCpaCBvZ2pa/d1.csv")
```

### Problem 1

```{r}
str(d1)
```

$H_0:$ Difference in mean ages equals zero.

$H_a:$ Difference in mean ages does not equal zero.

### Problem 2 and 3

```{r}
par(mfrow=c(1,2))
hist(d1$age, main = "Histogram of age", xlab = "Age")
hist(log10(d1$age), main = "Histogram of log age", xlab = "log(Age)")
```

 
 
 
A log transformation is not necessary in this case, methinks.
```{r}
## Make a QQ plot
qqnorm(d1$age)

## Add a theoretical line
qqline(d1$age, col = "red")

## Do a test to see if it's normally distributed
shapiro.test(d1$age)
```


### Problem 4

```{r}
control.age <- d1$age[d1$treatment == "control"]
vaccine.age <- d1$age[d1$treatment == "vaccinated"]

var.test(control.age, vaccine.age)

t.test(control.age, vaccine.age, var.equal = TRUE)

```

Conclusion: the mean difference between ages is 0.

### Problem 5

$H_0$ Mean HPV score in treatment = Mean HPV score in control
$H_a$ Mean HPV score in treatment does not equal Mean HPV score in control

### Problem 6 - 8

```{r}
hist(d1$score)
par(mfrow=c(1,2))
hist(d1$score, main = "Histogram of score", xlab = "score")
hist(log10(d1$score), main = "Histogram of log score", xlab = "log(score)")
```



The log of score looks normally distributed


```{r}
## predict that the log of score is normally distributed
logScore <- log(d1$score)

## Make a QQ plot
qqnorm(logScore)

## Add a theoretical line
qqline(logScore, col = "red")

## Do a test to see if it's normally distributed
shapiro.test(logScore)
```

### Problem 9

Seperate data into vaccine and control for score
```{r}
vaccineScore <- log(d1$score[d1$treatment == "vaccinated"])
controlScore <- log(d1$score[d1$treatment == "control"])

vaccineScoreRaw <- d1$score[d1$treatment == "vaccinated"]
controlScoreRaw <- d1$score[d1$treatment == "control"]
```

```{r}
var.test(vaccineScore, controlScore)
```


```{r}
t.test(vaccineScore, controlScore, var.equal = TRUE)
```


### Problem 10


```{r}
meanC <- mean(controlScoreRaw)
meanV <- mean(vaccineScoreRaw)

se <- function(x) sqrt(var(x)/length(x))

seC <- se(controlScoreRaw)
seV <- se(vaccineScoreRaw)

conInt <- function(x) c(mean(x) - (qnorm(0.975) * se(x)), mean(x) + (qnorm(0.975) * se(x)))

# Confidence interval for control
controlInt <- conInt(controlScoreRaw)

# For vaccine
vaccineInt <- conInt(vaccineScoreRaw)

my_bar <- barplot(c(meanC, meanV), xlab = "HPV susceptibility", names.arg = c("Control", "Vaccine"), ylim = c(0, 20))

arrows(x0 = my_bar, y0 = c(controlInt[1], vaccineInt[1]) , x1 = my_bar, y1 = c(controlInt[2], vaccineInt[2]), angle = 90, code = 3)

text(x = my_bar, y = c(controlInt[2], vaccineInt[2]), c("a", "b"), pos = 3)
```

### Problem 11

I want to approve this vaccine! I think it works.

### Problem 12

```{r}
meanV/meanC
meanC/meanV
```


### Problem 13

```{r}
# Download the data
d2 <- read.csv("https://nofile.io/f/KcWTFo0Rnst/d2.csv")
```

### Problem 14

```{r}
head(d2)
```


```{r}
scores <- aggregate(d2$score, FUN = mean, by = list(d2$trial, d2$treatment))
colnames(scores) <- c("Trial", "Treatment", "Mean Score")

scores
```

### Problem 15

```{r}
ratios <- matrix(data = NA, nrow = 10, ncol = 2)

for (i in 1:10) {
  table <- scores[scores["Trial"] == i]
  ## Control divided by vaccinated
  ratios[i, 1] <- i
  ratios[i, 2] <- as.numeric(table[5])/as.numeric(table[6])
}

colnames(ratios) <- c("Trial", "Ratio of Control to Vaccine")
rownames(ratios) <- 1:10
ratios
```


### Problem 16

```{r}
means <- mean(ratios[,2])

t.test(ratios[,2], mu = 3.01)
t.test(ratios[,2], mu = 3.2)
```

They're being a little misleading, but they're not lying (according to the T test). They also got the same values that we got!


Task 2
--------------

### Problem 1

```{r}
# Download the data
d3 <- read.csv("https://nofile.io/f/v8UMeTOzDq4/d3.csv")
```



### Problem 2

```{r}
str(d3)
```

```{r}
plot(d3$autism ~ d3$year, col = "purple", pch = 19, ylim = c(0, 20), ylab = "Incidence", xlab = "Year")
points(d3$vaccine ~ d3$year, col = "pink", pch = 19)
legend("topright", c("Autism", "Vaccines"), pch = c(19, 19), col = c("purple", "pink"))

```

The graph shows that austism and vaccines have both increased in the time between 1950 and 2000. This is NOT evidence of one causing the other, however.

### Problem 3

If I had the time and funding, I would create a longitudinal study. I would track mothers into groups controled for demographics and the like. After the mothers decided to either vaccine or not vaccine their children, I would try to pair children from similar demographs who either did or did not get vaccinated, and then compare autism rates.

### Problem 4

```{r}
# Download the data
d4 <- read.csv("https://nofile.io/f/EvDzvX7JYCf/d4.csv")
```

### Problem 5

```{r}
vac <- subset(d4, treatment == "vaccine")
control <- subset(d4, treatment == "control")
```






```{r}

x <- download.file("https://github.com/julianstanley/Biostats/blob/master", "d4.csv", method = "curl")

x
```






